Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-04 paper/paper.tex
tweaks to the introduction

The \gaia\ Mission (\citealt{gaia16}) will deliver more than a billion stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be both precise and purely astrometric.
\gaia\ will use astrometric parallax to determine the distances of the more precise
stars, and then calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
parallax estimates to less precise stars.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

However, the use of physical models for distance estimation is not necessary.
It is possible to build a stellar photometric model from the data themselves,
because there are stars at all luminosities with useful parallax information,.
One challenge is that different stars are observed at different levels of parallax precision,
so it requires relatively sophisticated technology to build this data-driven model
using all the data available, fairly and responsibly.
Luckily, we have this technology!

@@

We then use the novel \xd\ output---the deconvolved \cmd---as
a prior for use in inference of individual-star parallaxes.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is technically using the data twice and is therefore only approximately valid.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied\footnote{For a start, try \url{https://en.wikipedia.org/wiki/Empirical_Bayes_method}.}.

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to a related
project \citep{leistedtHogg2017}, in which we use a slightly less appropriate
model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

This project is fundamentally a demonstration of concept:
We are only using a subset of the ``small'' (relative to the expected final \gaia\ data set)
\tgas\ Catalog.
We are performing only an approximation to full Bayesian hierarchical inference.
We have to use photometry that is ground-based, rather than the full-precision,
space-based photometry that the \gaia\ Mission will deliver.
However, as we show below, we get very good results in terms of parallax precision.
It is promising that we will be able to obtain even better results
in later \gaia\ data releases, and eventually put distances on $>10^9$
\gaia\ stars using photometric parallax methodologies but without any
commitment to---or even use of---physical models of stars or the Milky Way.

>>>>

The \gaia\ Mission (\citealt{gaia16}) will soon deliver more than a billion
stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be both precise and purely astrometric.
\gaia\ will use astrometric parallaxes to determine the distances of the more
precise stars, and then calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
parallax estimates to less precise stars.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

However, the use of physical models for distance estimation is not necessary.
It is possible to build a stellar photometric model from the data themselves,
because there are stars at all luminosities with useful parallax information.
One challenge is that different stars are observed at different levels of parallax precision,
so it requires relatively sophisticated technology to build this data-driven model
using all of the data available, fairly and responsibly.

@@

We then use the novel \xd\ output---the deconvolved \cmd---as
a prior for use in inference of individual-star parallaxes.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is technically using the data twice and is therefore only approximately valid.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied.\footnote{For a primer on ``empirical Bayes'', see \url{https://en.wikipedia.org/wiki/Empirical_Bayes_method}.}

A probabilistically valid approach that wouldn't re-use the data is to perform
a full hierarchical Bayesian inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to a related
project \citep{leistedtHogg2017}, in which we use a slightly less appropriate
model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

This project is fundamentally a demonstration of concept:
We are only using a subset of the ``small'' (relative to the expected final \gaia\ data set)
\tgas\ Catalog.
We are performing only an approximation to full Bayesian hierarchical inference.
We have to use photometry that is ground-based, rather than the full-precision,
space-based photometry that the \gaia\ Mission will deliver.
However, as we show below, we get very good results in terms of parallax precision.
It is promising that we will be able to obtain even better results
using later \gaia\ data releases, and eventually infer distances for $>10^9$
\gaia\ stars using photometric parallax methodologies but without any
commitment to---or even use of---physical models of stars or the Milky Way.
